---
title: "Microclimate Model Correction Workflow with microclCorr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Microclimate Model Correction Workflow with microclCorr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/vignette-",
  out.width = "100%"
)
```

## Introduction

Physical microclimate models such as **NicheMapR** calculate microenvironmental conditions (temperature, humidity, solar radiation) from first physical principles, including terrain topography, radiative transfer, and convective heat balance. While these mechanistic models provide broad spatiotemporal predictions, local microtopography, fine-scale shading, and boundary-layer dynamics often introduce systematic prediction biases.

`microclCorr` provides a machine learning post-processing framework to correct these systematic residuals using in-situ temperature logger data. Rather than replacing the physical model, `microclCorr` models the residual difference:

$$\text{residual} = T_{\text{measured}} - T_{\text{predicted}}$$

Once the machine learning model estimates the residual ($\widehat{\text{residual}}$), the final corrected temperature prediction is:

$$T_{\text{corrected}} = T_{\text{predicted}} + \widehat{\text{residual}}$$

The package supports two machine learning paradigms:
- **Random Forest (RF)**: Robust, tabular ensemble learning that handles nonlinear feature interactions with minimal tuning.
- **Long Short-Term Memory (LSTM)**: Recurrent neural network that captures temporal sequence dependencies and thermal inertia over sliding time windows.

---

## Workflow Overview

The correction pipeline consists of data preparation, feature engineering, model training and hyperparameter tuning, model evaluation, and deployment to new predictions:

```{r workflow-diagram, echo = FALSE, out.width = "95%", fig.align = "center", purl = FALSE}
img_path <- if (file.exists("workflow_combined.png")) "workflow_combined.png" else file.path("vignettes", "workflow_combined.png")
if (file.exists(img_path)) knitr::include_graphics(img_path)
```

---

## Pipeline Stages

### 1. Data Preparation and Ingestion

The training pipeline begins with time-aligned observations where physical model predictions and in-situ logger measurements have been paired. Columns typically include:

- `time`: Date-time timestamp
- `measured`: Logger measurement ($^\circ\text{C}$)
- `predicted`: Physical model prediction ($^\circ\text{C}$)
- `residual`: Measured temperature minus predicted temperature ($T_{\text{measured}} - T_{\text{predicted}}$)
- Environmental predictors (e.g. ambient air temperature `TAREF`, solar radiation `SOLR`, wind speed `VREF`, relative humidity `RH`)
- Optional grouping columns (e.g. categorical `microhabitat` or time series identifier `time_series_doc`)

`microclCorr` offers two ways to ingest and prepare datasets:

#### Option A: Ingest from an aligned CSV file (`load_prepared_csv_data`)
```r
library(microclCorr)

# Ingest prepared CSV from disk or package example data
csv_path <- get_example_data("Harod_dataset.csv")
data <- load_prepared_csv_data(
  path                       = csv_path,
  is_continuous_microhabitat = FALSE,
  datetime_format            = "%d/%m/%Y %H:%M",
  includes_index             = TRUE
)
```

#### Option B: Prepare an in-memory data.frame (`prepare_dataframe`)
If your dataset is already loaded into your R session (for instance, the bundled `microclimate_sample` dataset or a data frame constructed programmatically), use `prepare_dataframe()`:

```r
# Load bundled sample dataset
data(microclimate_sample)

# Parse datetime, encode microhabitat, and validate structure
data <- prepare_dataframe(
  df                         = microclimate_sample,
  is_continuous_microhabitat = FALSE,
  datetime_col               = "time",
  microhabitat_col           = "microhabitat"
)
```

---

### 2. Feature Engineering & Temporal Splitting

Cyclical temporal features (hour of day, and optionally month of year) are extracted using sine and cosine transformations:

```r
# Add cyclical diurnal features (Hour_sin, Hour_cos)
data <- add_cyclical_time(data, datetime_col = "time", add_month = FALSE)

# Automatically identify numeric environmental predictor columns
feature_cols <- get_feature_columns(data)
```

To avoid temporal data leakage (where random row-level splitting lets models interpolate between adjacent hours), `microclCorr` uses **blocked cross-validation**:

```r
# Block-temporal split: 70% train, 15% validation, 15% test in contiguous 2-day blocks
splits <- split_train_val_test(
  data       = data,
  train_pct  = 0.70,
  val_pct    = 0.15,
  block_days = 2,
  seed       = 42
)
```

For datasets containing multiple loggers, sites, or microhabitats, use `stratified_split_train_val_test()` to ensure balanced temporal representation across all groups.

---

### 3. Model Training

#### Random Forest
Random Forest fits an ensemble of regression trees to predict residuals from environmental features:

```r
rf_model <- train_rf(
  train_X        = splits$train[, feature_cols],
  train_y        = splits$train$residual,
  num_trees      = 500,
  tune           = TRUE,
  n_combinations = 10,
  val_X          = splits$val[, feature_cols],
  val_y          = splits$val$residual,
  seed           = 42
)
```

#### LSTM Neural Network

LSTM models capture temporal memory and thermal inertia. Because LSTM models rely on TensorFlow and Keras via `reticulate`, verify or configure the environment at the start of your session:

```r
# Check or configure Python/TensorFlow environment
setup_tensorflow()
```

Data are scaled and structured into fixed sliding sequence windows before model training:

```r
# Normalize features using min-max scaling fitted strictly on training data
scaled <- lstm_scaling(splits$train, splits$val, splits$test)

# Reshape continuous time series into sliding sequence windows (e.g. 4-hour lookback)
lstm_data <- lstm_specific_preprocessing(
  train        = scaled$train,
  val          = scaled$val,
  test         = scaled$test,
  window_size  = 4,
  ts_names_col = "time_series_doc"
)
```

##### Automated Hyperparameter Tuning (`lstm_hypertuning`)

Rather than relying on arbitrary neural architecture choices, `lstm_hypertuning()` performs random search across key structural and optimization parameters:

```r
# Search for optimal LSTM architecture over random trials
hpo <- lstm_hypertuning(
  train_X       = lstm_data$train_dict$X,
  train_y       = lstm_data$train_dict$y,
  val_X         = lstm_data$val_dict$X,
  val_y         = lstm_data$val_dict$y,
  n_trials      = 5,
  units_range   = c(32, 128),
  layers_range  = c(1, 2),
  dropout_range = c(0.0, 0.2),
  lr_range      = c(1e-4, 1e-2),
  epochs        = 40,
  batch_size    = 32,
  patience      = 8,
  seed          = 42
)

best_lstm <- hpo$model
```

Alternatively, train directly with chosen hyperparameters using `train_lstm()`.

---

### 4. Evaluation and Model Selection

Performance is quantified by comparing the Root Mean Square Error (RMSE) and coefficient of determination ($R^2$) of the raw physical model predictions against the ML-corrected predictions.

#### Aligning Test Sets (`align_test_sets`)
Because LSTM sequence windows omit the first $(window\_size - 1)$ hours of each time series, `align_test_sets()` filters the Random Forest test set to evaluate both models on identical timestamps:

```r
# Align tabular test set to match LSTM window endpoints
aligned_test_rf <- align_test_sets(
  test_dataset   = splits$test,
  lstm_test_dict = lstm_data$test_dict,
  ts_index_info  = lstm_data$index_info,
  site_name_col  = "time_series_doc",
  datetime_col   = "time"
)

# Evaluate Random Forest on aligned test set
metrics_rf <- evaluate_correction(
  model           = rf_model,
  X               = aligned_test_rf[, feature_cols],
  y               = aligned_test_rf$residual,
  base_prediction = aligned_test_rf$predicted,
  model_type      = "rf"
)

# Evaluate LSTM on test windows
metrics_lstm <- evaluate_correction(
  model           = best_lstm,
  X               = lstm_data$test_dict$X,
  y               = lstm_data$test_dict$y,
  base_prediction = lstm_data$test_dict$base_pred,
  model_type      = "lstm"
)

print(metrics_rf)
print(metrics_lstm)
```

---

### 5. Model Persistence (`save_correction_model` / `load_correction_model`)

Once trained, models can be bundled with their required metadata (feature columns, model type, and scaling parameters) into a single `.rds` file:

```r
# Save Random Forest model bundle
save_correction_model(
  model        = rf_model,
  feature_cols = feature_cols,
  path         = "best_rf_model.rds"
)

# Save LSTM bundle (automatically includes scaler and serializes neural network weights)
save_correction_model(
  model        = best_lstm,
  scaler       = scaled$scaler,
  feature_cols = feature_cols,
  path         = "best_lstm_model.rds"
)

# Load the saved model bundle in any new R session
bundle <- load_correction_model("best_rf_model.rds")
```

---

### 6. Predicting on New Data (`correct_predictions`)

The ultimate objective of `microclCorr` is to correct **new, unmeasured physical model predictions**—such as future weather forecasts, historical reconstructions, unmeasured seasons, or new microhabitats where no field loggers exist.

In this scenario:
- **`new_data`** contains the mechanistic model predictions (e.g. `predicted` column from NicheMapR) and environmental forcing features (`TAREF`, `SOLR`, etc.).
- Ground-truth logger measurements (`measured`) and target residuals (`residual`) are **not available** and not needed.

The function `correct_predictions()` applies the trained model, estimates the expected residual correction, and returns the final corrected microclimate values:

$$\text{corrected\_prediction} = \text{base\_prediction} + \text{correction}$$

```r
# Apply model to new physical model predictions
corrected_df <- correct_predictions(
  model        = bundle$model,
  new_data     = new_physical_data,
  model_type   = bundle$model_type,
  feature_cols = bundle$feature_cols,
  scaler       = bundle$scaler # required for LSTM, ignored for RF
)
```

The output is a clean data.frame with:
- `datetime`: Timestamps of each prediction point
- `base_prediction`: Uncorrected physical model predictions
- `correction`: Estimated residual correction ($\widehat{\text{residual}}$)
- `corrected_prediction`: Final corrected temperature ($T_{\text{predicted}} + \widehat{\text{residual}}$)

---

## Reproducible End-to-End Example

The following self-contained example runs the complete pipeline on the bundled `microclimate_sample` dataset, demonstrating model training, evaluation, saving/loading, and **predicting on new unmeasured data**.

```{r reproducible-example}
library(microclCorr)

# ── 1. Load Bundled Empirical Data ───────────────────────────────────────────
data(microclimate_sample)

# Prepare dataset (parses datetime, validates columns, one-hot encodes microhabitat)
prepared <- prepare_dataframe(microclimate_sample)

# Define predictor feature columns
features <- c("TAREF", "RH", "VREF", "SOLR", "TSKYC", "DEW",
              "Hour_sin", "Hour_cos",
              "microhabitat_air", "microhabitat_shade", "microhabitat_sun")

# ── 2. Blocked Temporal Split ────────────────────────────────────────────────
# 70% training, 15% validation, 15% testing in 2-day contiguous blocks
splits <- split_train_val_test(
  prepared,
  train_pct  = 0.70,
  val_pct    = 0.15,
  block_days = 2,
  seed       = 42
)

# ── 3. Train Random Forest Model ─────────────────────────────────────────────
rf_model <- train_rf(
  train_X   = splits$train[, features],
  train_y   = splits$train$residual,
  num_trees = 100,
  tune      = FALSE,
  seed      = 42
)

# ── 4. Evaluate Correction Performance on Test Set ───────────────────────────
eval_results <- evaluate_correction(
  model           = rf_model,
  X               = splits$test[, features],
  y               = splits$test$residual,
  base_prediction = splits$test$predicted,
  model_type      = "rf"
)

cat(sprintf("Baseline Uncorrected RMSE : %.2f deg C (R^2 = %.3f)\n",
            eval_results$rmse_base, eval_results$r2_base))
cat(sprintf("ML-Corrected RMSE         : %.2f deg C (R^2 = %.3f)\n",
            eval_results$rmse_corr, eval_results$r2_corr))
cat(sprintf("RMSE Error Reduction      : %.1f%%\n",
            100 * (1 - eval_results$rmse_corr / eval_results$rmse_base)))

# ── 5. Save and Reload Model Bundle ──────────────────────────────────────────
tmp_model_file <- tempfile(fileext = ".rds")
save_correction_model(
  model        = rf_model,
  feature_cols = features,
  path         = tmp_model_file
)

# Load the saved model bundle
bundle <- load_correction_model(tmp_model_file)
unlink(tmp_model_file)

# ── 6. Predicting on New Unmeasured Data ─────────────────────────────────────
# Simulate a new dataset of physical model predictions where no field loggers exist.
# Notice that ground-truth measurements (residual, logger columns) are omitted:
new_physical_data <- splits$test
new_physical_data$residual   <- NULL  # Residuals are unknown in real deployment!
new_physical_data$sun_temp   <- NULL
new_physical_data$shade_temp <- NULL
new_physical_data$air_temp   <- NULL

# Apply the correction model to the new physical predictions
corrected_predictions <- correct_predictions(
  model        = bundle$model,
  new_data     = new_physical_data,
  model_type   = bundle$model_type,
  feature_cols = bundle$feature_cols
)

# Inspect the corrected output
head(corrected_predictions)
```

Notice that `correct_predictions()` seamlessly returns the original mechanistic baseline (`base_prediction`), the estimated correction (`correction`), and the final corrected value (`corrected_prediction`).

---

### Extending to LSTM Models

For time series exhibiting significant thermal inertia, an LSTM recurrent neural network can be trained and deployed using the exact same workflow:

```{r lstm-pipeline, eval = FALSE}
# 0. Configure Python and TensorFlow environment
setup_tensorflow()

# 1. Feature normalization (min-max scaling fit on training set)
scaled <- lstm_scaling(splits$train, splits$val, splits$test)

# 2. Reshape into sliding sequence windows (e.g. 4-hour temporal lookback)
lstm_data <- lstm_specific_preprocessing(
  train        = scaled$train,
  val          = scaled$val,
  test         = scaled$test,
  window_size  = 4,
  ts_names_col = "time_series_doc"
)

# 3. Train LSTM model (or use lstm_hypertuning for automated tuning)
lstm_model <- train_lstm(
  train_X    = lstm_data$train_dict$X,
  train_y    = lstm_data$train_dict$y,
  val_X      = lstm_data$val_dict$X,
  val_y      = lstm_data$val_dict$y,
  n_units    = 64,
  n_layers   = 2,
  dropout    = 0.1,
  lr         = 0.001,
  epochs     = 50,
  batch_size = 32,
  seed       = 42
)

# 4. Save LSTM bundle (includes scaler parameters)
save_correction_model(
  model        = lstm_model,
  scaler       = scaled$scaler,
  feature_cols = features,
  path         = "best_lstm_model.rds"
)

# 5. Predict on new unmeasured data using the saved scaler
loaded_lstm <- load_correction_model("best_lstm_model.rds")

corrected_lstm <- correct_predictions(
  model        = loaded_lstm$model,
  new_data     = new_physical_data,
  model_type   = "lstm",
  feature_cols = loaded_lstm$feature_cols,
  scaler       = loaded_lstm$scaler,
  window_size  = 4,
  datetime_col = "time"
)

head(corrected_lstm)
```
